```{r chunkSetup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(SASxport)
require(DiagrammeR)
require(DiagrammeRsvg)
require(rsvg)
library(magrittr)
library(svglite)
library(png)
require(nhanesA)
require(survey)
require(Publish)
require(knitr)
require(jtools)
require(interactions)
require(interactionR)
cache.ON = TRUE
```

# Interaction

We will be using three measures of effect to explain the concept and necessary implementation ideas


## Modelling to obtain OR

 - odds ratio (OR),  Multiplicative scale
    - Logit[Pr(Y=1)] = $\alpha_0$ + $\alpha_A$ A + $\alpha_B$ B + $\alpha_{AB}$ (A $\times$ B)
    - Condition for presence of interaction ("departure from multiplicativity")
        - OR10 $\times$ OR01 $\ne$ OR11
        - $\exp(\alpha_A)$ $\times$ $\exp(\alpha_B)$ $\ne$ $\exp(\alpha_A+ \alpha_B + \alpha_{AB})$
    - synergism / super-multiplicative
        - OR11 > OR10 $\times$ OR01
    - antagonism / sub-multiplicative
        - OR11 < OR10 $\times$ OR01
        
*Note*: 
The above assumes that both exposures are risk factors for the outcome (OR > 1), not protective factors. If protective, estimates of RERI and AP (defined later) will be invalid, although the estimate of SI is not affected by this condition.

### Base 

#### no alcohol, no smoking for OR(smk1 on outcome [alc1==0] and OR(alc1 on outcome [smk1==0]

Model-based SE estimates from logistic regression model [@naimi2020estimating]:

```{r}
require(interactionR)
data(OCdata)
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "0")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "0")
fit.or.int11 <- glm(oc ~ alc + smk + alc:smk, 
                    family = binomial(link = 'logit'), 
                    data = Obs.Data)
require(jtools)
results.or.model11 <- summ(fit.or.int11, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = TRUE, 
                           confint = TRUE)
results.or.model11
```

### Estimates of different ORs

<!----
###### Base 

- $\alpha_0$ = `r round(as.numeric(fit.or.int11$coefficients['(Intercept)']),2)`
- $\exp(\alpha_0)$ = `r round(exp(as.numeric(fit.or.int11$coefficients['(Intercept)'])),2)`
--->


##### Alcohol

```{r}
## OR00 = 1
OR10 <- exp(sum(summary(fit.or.int11)$coef[c('alc1'),'Estimate']))
OR10 ## OR_A=1
```

- $\alpha_A$ = `r round(as.numeric(fit.or.int11$coefficients['alc1']),2)`
- OR_{A=1} = OR10 = $\exp(\alpha_A)$ = `r round(exp(as.numeric(fit.or.int11$coefficients['alc1'])),2)`

##### Smoking

```{r}
OR01 <- exp(sum(summary(fit.or.int11)$coef[c('smk1'),'Estimate']))
OR01 ## OR_B=1
```

- $\alpha_B$ = `r round(as.numeric(fit.or.int11$coefficients['smk1']),2)`
- OR_{B=1} = OR01 = $\exp(\alpha_B)$ = `r round(exp(as.numeric(fit.or.int11$coefficients['smk1'])),2)`

##### Both alcohol and smoking

```{r}
OR11 <- exp(sum(summary(fit.or.int11)$coef[c('alc1','smk1','alc1:smk1'),'Estimate'])) 
OR11 ##OR_A=1,B=1
```

- $\alpha_{AB}$ = `r round(as.numeric(fit.or.int11$coefficients['alc1:smk1']),2)`
- OR_{A=1,B=1} = OR11 = `r round(exp(sum(summary(fit.or.int11)$coef[c('alc1','smk1','alc1:smk1'),'Estimate'])),2)`

##### Joint effect?

Is OR11 $\ne$ OR10 * OR01

```{r}
OR10 * OR01
```

##### Additive scales

- Relative excess risk due to interaction (RERI)
  - ranges from 0 to $\infty$
  - 0 means no interaction
  - +ve means positive interaction 
- Proportion of outcome among those with both exposures that is attributable to the interaction (AP)
  - ranges from -1 to 1
  - 0 means no interaction 
- Synergy index (SI) 
  - ratio of the combined effects and the individual effects
  - ranges from 0 to $\infty$
  - greater than 1 means positive interaction


```{r}
RERI <- OR11 - OR10 - OR01 + 1
RERI
AP <- RERI / OR11
AP
SI <- (OR11 - 1)/ (OR10 - 1 + OR01 - 1)
SI
```

##### Recoding to obtain different combinations

###### Base: alcohol drinker, no smoking for OR(smk1 on outcome [alc1==1]

```{r}
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "0")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "1")
fit.or.int01 <- glm(oc ~ alc + smk + alc:smk, 
                    family = binomial(link = 'logit'), 
                    data = Obs.Data)
results.or.model01 <- summ(fit.or.int01, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = TRUE, 
                           confint = TRUE)
results.or.model01
```

###### Base: Smoker, no alcohol for OR(alc1 on outcome [smk1==1]

```{r}
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "1")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "0")
fit.or.int10 <- glm(oc ~ alc + smk + alc:smk, 
                    family = binomial(link = 'logit'), 
                    data = Obs.Data)
results.or.model10 <- summ(fit.or.int10, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = TRUE,  
                           confint = TRUE)
results.or.model10

OR10b <- exp(sum(summary(fit.or.int10)$coef[c('alc1'),'Estimate']))
OR10b
```

#### Multiplicative scale

```{r}
Multiplicative.scale <- exp(sum(summary(fit.or.int11)$coef[c('alc1:smk1'),'Estimate'])) 
Multiplicative.scale
```

- Multiplicative.scale = $\exp(\alpha_{AB})$ = `r round(exp(as.numeric(fit.or.int11$coefficients['alc1:smk1'])),2)`

```{r}
# estimates of alcohol after changing smoking ref levels
OR10b / OR10
```

### Reporting guideline

Ref: @knol2012recommendations, @alli2021interactionr

```{r}
int.object <- interactionR(fit.or.int11, 
                            exposure_names = c("alc1", "smk1"), 
                            ci.type = "mover", ci.level = 0.95, 
                            em=FALSE, recode = FALSE)
kable(int.object$dframe[,1:4], digits = 2)
```

### Other available functions

Ref: @mathur2018r

```{r}
## Mathur MB & VanderWeele TJ (2018)
source("Additive interactions function.R") 
additive_interactions(fit.or.int11)
```

### Interaction definitions for OR

| Groups and Conditions 	| Risk 	| Odds Ratio |
|---	|---	|---	|
| Baseline (exposed to none) 	| $R_{A = 0, B = 0} =  P[Y^{a=0, b=0} | L]$ 	| Reference | 
| Exposed to $A$ only 	| $R_{A = 1, B = 0} =  P[Y^{a=1, b=0} | L]$ 	| $OR_{A = 1} = [R_{A = 1, B = 0} / (1-R_{A = 1, B = 0})] / [R_{A = 0, B = 0} / (1 - R_{A = 0, B = 0})]$ |
| Exposed to $B$ only 	| $R_{A = 0, B = 1} =  P[Y^{a=0, b=1} | L]$ 	| $OR_{B = 1} = [R_{A = 0, B = 1} / (1-R_{A = 0, B = 1})] / [R_{A = 0, B = 0} / (1 - R_{A = 0, B = 0})]$ |
| Exposed to both $A$ and $B$ 	| $R_{A = 1, B = 1} =  P[Y^{a=1, b=1} | L]$ 	| $OR_{A = 1, B = 1} = [R_{A = 1, B = 1} / (1-R_{A = 1, B = 1})] / [R_{A = 0, B = 0} / (1 - R_{A = 0, B = 0})]$ |
| **Condition for interaction** 	 	|  | $OR_{A = 1, B = 1} \ne OR_{A = 1} \times OR_{B = 1}$	|
| **Synergism** 	 	||   $OR_{A = 1, B = 1} > OR_{A = 1} \times OR_{B = 1}$	|
| **Antagonism** 	 	||  $OR_{A = 1, B = 1} < OR_{A = 1} \times OR_{B = 1}$	|
Table: \label{tab:def} Summary of interaction definition for OR.


## Modelling to obtain RR

 - risk ratio (RR), Multiplicative scale
    - Very similar ideas as in OR
    
```{r}
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "0")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "0")
fit.rr.int11 <- glm(oc ~ alc + smk + alc:smk, 
                    family = poisson(link = 'log'), 
                    data = Obs.Data)
results.rr.model11 <- summ(fit.rr.int11, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = TRUE, 
                           robust = "HC1", 
                           confint = TRUE)
results.rr.model11
```

Ref: @naimi2020estimating

Alternate formulation using log-binomial is possible (using model based SE, but confidence intervals somewhat differ from those reported from poisson(log)):

```{r}
fit.rr.int11lb <- glm(oc ~ alc + smk + alc:smk, 
                      family = binomial(link = 'log'), 
                      data = Obs.Data)
results.rr.model11lb <- summ(fit.rr.int11lb, 
                             model.info = FALSE, 
                             model.fit = FALSE, 
                             exp = TRUE, 
                             confint = TRUE)
results.rr.model11lb
```

### Estimates of different RRs

#### Alcohol

```{r}
## RR00 = 1
RR10 <- exp(sum(summary(fit.rr.int11)$coef[c('alc1'),'Estimate']))
RR10 ## RR_A=1
```

- $\alpha_A$ = `r round(as.numeric(fit.rr.int11$coefficients['alc1']),2)`
- RR_{A=1} = RR10 = $\exp(\alpha_A)$ = `r round(exp(as.numeric(fit.rr.int11$coefficients['alc1'])),2)`

#### Smoking

```{r}
RR01 <- exp(sum(summary(fit.rr.int11)$coef[c('smk1'),'Estimate']))
RR01 ## RR_B=1
```

- $\alpha_B$ = `r round(as.numeric(fit.rr.int11$coefficients['smk1']),2)`
- RR_{B=1} = RR01 = $\exp(\alpha_B)$ = `r round(exp(as.numeric(fit.rr.int11$coefficients['smk1'])),2)`

#### Both alcohol and smoking

```{r}
RR11 <- exp(sum(summary(fit.rr.int11)$coef[c('alc1','smk1','alc1:smk1'),'Estimate'])) 
RR11 ##RR_A=1,B=1
```

- $\alpha_{AB}$ = `r round(as.numeric(fit.rr.int11$coefficients['alc1:smk1']),2)`
- RR_{A=1,B=1} = RR11 = `r round(exp(sum(summary(fit.rr.int11)$coef[c('alc1','smk1','alc1:smk1'),'Estimate'])),2)`

#### Joint effect?

Is RR11 $\ne$ RR10 * RR01

```{r}
RR10 * RR01
```

#### Multiplicative scale

```{r}
Multiplicative.scale <- exp(sum(summary(fit.rr.int11)$coef[c('alc1:smk1'),'Estimate'])) 
Multiplicative.scale
```

- Multiplicative.scale = $\exp(\alpha_{AB})$ = `r round(exp(as.numeric(fit.rr.int11$coefficients['alc1:smk1'])),2)`

#### Additive scales

```{r}
RERI <- RR11 - RR10 - RR01 + 1
RERI
AP <- RERI / RR11
AP
SI <- (RR11 - 1)/ (RR10 - 1 + RR01 - 1)
SI
```


#### Reporting guideline

```{r}
int.object <- interactionR(fit.rr.int11, 
                            exposure_names = c("alc1", "smk1"), 
                            ci.type = "mover", ci.level = 0.95, 
                            em=FALSE, recode = FALSE)
kable(int.object$dframe[,1:4], digits = 2)
```

Ref: @alli2021interactionr

#### Interaction definitions for RR

| Groups and Conditions 	| Risk 	| Risk Ratio | 
|---	|---	|---	|---	| ---	|
| Baseline (exposed to none) 	| $R_{A = 0, B = 0} =  P[Y^{a=0, b=0} | L]$ 	| Reference	|   
| Exposed to $A$ only 	| $R_{A = 1, B = 0} =  P[Y^{a=1, b=0} | L]$ 	|  $RR_{A = 1} = R_{A = 1, B = 0} / R_{A = 0, B = 0}$	| 
| Exposed to $B$ only 	| $R_{A = 0, B = 1} =  P[Y^{a=0, b=1} | L]$ 	|  $RR_{B = 1} = R_{A = 0, B = 1} / R_{A = 0, B = 0}$	| 
| Exposed to both $A$ and $B$ 	| $R_{A = 1, B = 1} =  P[Y^{a=1, b=1} | L]$ 	|  $RR_{A = 1, B = 1} = R_{A = 1, B = 1} / R_{A = 0, B = 0}$	| 
| **Condition for interaction** 	 	||  $RR_{A = 1, B = 1} \ne RR_{A = 1} \times RR_{B = 1}$ 	
| **Synergism** 	 	||   $RR_{A = 1, B = 1} > RR_{A = 1} \times RR_{B = 1}$ 	|
| **Antagonism** 	 	||   $RR_{A = 1, B = 1} < RR_{A = 1} \times RR_{B = 1}$ 	|
Table: \label{tab:def} Summary of interaction definition for RR.


## Modelling to obtain RD

 - risk difference (RD), Additive scale
    - risk(Y) = $\alpha_0$ + $\alpha_A$ A + $\alpha_B$ B + $\alpha_{AB}$ (A $\times$ B)
        - RD10 $\times$ RD01 $\ne$ RD11
        - $\alpha_A$ + $\alpha_B$ $\ne$ ($\alpha_A$ + $\alpha_B$ + $\alpha_{AB}$)
    - synergism / super-additive
        - RD11 > RD10 + RD01
    - antagonism / sub-additive
        - RD11 < RD10 + RD01    


```{r}
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "0")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "0")
fit.rd.int11 <- glm(oc ~ alc + smk + alc:smk, 
                    family = gaussian(link = 'identity'), 
                    data = Obs.Data)
results.rd.model11 <- summ(fit.rd.int11, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = FALSE, 
                           robust = "HC1", 
                           confint = TRUE)
results.rd.model11
```

Ref: @naimi2020estimating


Alternate formulation using binomial(identity) is possible (using model based SE, but confidence intervals slightly differ from those reported from gaussian(identity)):

```{r}
fit.rd.int11bi <- glm(oc ~ alc + smk + alc:smk, 
                      family = binomial(link = 'identity'), 
                      data = Obs.Data)
results.rd.model11bi <- summ(fit.rd.int11bi, 
                             model.info = FALSE, 
                             model.fit = FALSE, 
                             exp = FALSE, 
                             confint = TRUE)
results.rd.model11bi
```

### Base: Smoker, no alcohol for RD(alc1 on outcome [smk1==1]
Alternate combination of smoking:

```{r}
Obs.Data <- OCdata
Obs.Data$smk <- as.factor(Obs.Data$smk)
Obs.Data$smk <- relevel(Obs.Data$smk, ref = "1")
Obs.Data$alc <- as.factor(Obs.Data$alc)
Obs.Data$alc <- relevel(Obs.Data$alc, ref = "0")
fit.rd.int01 <- glm(oc ~ alc + smk + alc:smk, 
                    family = gaussian(link = 'identity'), 
                    data = Obs.Data)
results.rd.model01 <- summ(fit.rd.int01, 
                           model.info = FALSE, 
                           model.fit = FALSE, 
                           exp = FALSE, 
                           robust = "HC1", 
                           confint = TRUE)
results.rd.model01
```


### Estimates of different RDs

#### Alcohol

```{r}
## RD00 = 0
RD10 <- sum(summary(fit.rd.int11)$coef[c('alc1'),'Estimate'])
RD10 ## RD_A=1
```

- RD_{A=1} = RD10 = $\alpha_A$ = `r round(as.numeric(fit.rd.int11$coefficients['alc1']),2)`

From alternate combination of smoking

```{r}
RD10b <- sum(summary(fit.rd.int01)$coef[c('alc1'),'Estimate'])
RD10b ## RD_A=1
```


#### Smoking

```{r}
RD01 <- sum(summary(fit.rd.int11)$coef[c('smk1'),'Estimate'])
RD01 ## RD_B=1
```

- RD_{B=1} = RD01 = $\alpha_B$ = `r round(as.numeric(fit.rd.int11$coefficients['smk1']),2)`


#### Both alcohol and smoking

```{r}
RD11 <- sum(summary(fit.rd.int11)$coef[c('alc1','smk1','alc1:smk1'),'Estimate'])
RD11 ##RD_A=1,B=1
```

- RD_{A=1,B=1} = RD11 = $\alpha_{AB}$ = `r round(as.numeric(fit.rd.int11$coefficients['alc1:smk1']),2)`

#### Joint effect?

Is RD11 $\ne$ RD10 + RD01

```{r}
RD10 + RD01
```

#### Additive scale (interaction contrast)

```{r}
IC <- RD10b - RD10
IC
```

Current version of `interactionR` do not support RD, although the documentation says so: @alli2021interactionr

### Interaction definitions for RD

| Groups and Conditions 	| Risk 	| Risk Difference 	| 
|---	|---	|---	|---	| ---	|
| Baseline (exposed to none) 	| $R_{A = 0, B = 0} =  P[Y^{a=0, b=0} | L]$ 	| Reference	|  
| Exposed to $A$ only 	| $R_{A = 1, B = 0} =  P[Y^{a=1, b=0} | L]$ 	| $RD_{A = 1} = R_{A = 1, B = 0} - R_{A = 0, B = 0}$	| 
| Exposed to $B$ only 	| $R_{A = 0, B = 1} =  P[Y^{a=0, b=1} | L]$ 	| $RD_{B = 1} = R_{A = 0, B = 1} - R_{A = 0, B = 0}$	| 
| Exposed to both $A$ and $B$ 	| $R_{A = 1, B = 1} =  P[Y^{a=1, b=1} | L]$ 	| $RD_{A = 1, B = 1} = R_{A = 1, B = 1} - R_{A = 0, B = 0}$	| 
| **Condition for interaction** 	 	||  $RD_{A = 1, B = 1} \ne RD_{A = 1} + RD_{B = 1}$	| 
| **Synergism** 	 	||  $RD_{A = 1, B = 1} > RD_{A = 1} + RD_{B = 1}$	| 
| **Antagonism** 	 	||  $RD_{A = 1, B = 1} < RD_{A = 1} + RD_{B = 1}$	| 
Table: \label{tab:def} Summary of interaction definition for RD.


## Reporting guidelines

- How to determine whether interaction should be reported in additive or multiplicative scale?
  - Not via empirical assessment (analysis of data at hand)
  - Study objectives should be clear
  - data generating mechanism should be considered carefully
- What if results differ?
  - Very possible to have additive interaction present, but not at a multiplicative scale and vice versa
  - When conclusions differ (interaction is present in additive, but absent in multiplicative or vice versa), both should be reported.
  - When both are concluding the same (interaction is present in both scales or absent), questions related to magnitude remains


